---
title: "Data loading and harmonizing"
subtitle: "Preparing CCHS data for the Smoking Histories Model"
author: "CSHM Development Team"
date: "Last Updated: `r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load required packages
library(dplyr)
library(ggplot2)
# If you have installed the cshgm package, uncomment this:
# library(cshm)
```

# Introduction

This guide focuses on the initial steps of the CSHM workflow: loading Canadian Community Health Survey (CCHS) data and harmonizing variables across survey cycles. These steps are critical for ensuring consistent analysis across different time periods.

## Purpose of data harmonization

Variable names, coding, and survey design have changed across CCHS cycles. Harmonization ensures that:

1. Variables have consistent names across cycles
2. Response codes are standardized
3. Derived variables are calculated consistently
4. Missing values are handled appropriately

# Required data sources

## Canadian Community Health Survey (CCHS)

The CCHS is the primary data source for the smoking history model. We typically use multiple cycles:

- CCHS 2001 (Cycle 1.1)
- CCHS 2003 (Cycle 2.1)
- CCHS 2005 (Cycle 3.1)
- CCHS 2007-2008
- CCHS 2009-2010
- CCHS 2011-2012
- CCHS 2013-2014
- CCHS 2015-2016
- CCHS 2017-2018

## Accessing CCHS data

CCHS data can be accessed through:

1. **Statistics Canada Research Data Centres (RDCs)**: For researchers who need access to the detailed microdata
2. **Public Use Microdata Files (PUMFs)**: Available through academic institutions
3. **Sample files**: For demonstration purposes, sample datasets are included with this package

# Loading CCHS data

## Step 1: Set up your environment

First, set up your environment with the necessary packages:

```{r load-packages, eval=FALSE}
# Load required packages
library(cshm)
library(dplyr)
library(cchsflow)  # For harmonizing CCHS variables
library(haven)     # For reading SAS or SPSS files

# Configure paths using the config package
library(config)
cfg <- config::get()
data_path <- cfg$data_path
```

## Step 2: Load individual CCHS cycles

Load each CCHS cycle file. The format will depend on how your data is stored:

```{r load-cycles, eval=FALSE}
# Example: Loading RData files
cchs2001 <- readRDS(file.path(data_path, "cchs2001_p.RData"))
cchs2003 <- readRDS(file.path(data_path, "cchs2003_p.RData"))
cchs2005 <- readRDS(file.path(data_path, "cchs2005_p.RData"))
cchs2007_2008 <- readRDS(file.path(data_path, "cchs2007_2008_p.RData"))
cchs2009_2010 <- readRDS(file.path(data_path, "cchs2009_2010_p.RData"))
cchs2011_2012 <- readRDS(file.path(data_path, "cchs2012_p.RData"))
cchs2013_2014 <- readRDS(file.path(data_path, "cchs2013_2014_p.RData"))

# For SAS or SPSS files, you would use:
# cchs2001 <- haven::read_sas(file.path(data_path, "cchs2001.sas7bdat"))
# or
# cchs2001 <- haven::read_spss(file.path(data_path, "cchs2001.sav"))
```

## Step 3: Data preparation

Before harmonizing, prepare the data by adding cycle-specific identifiers:

```{r prepare-data, eval=FALSE}
# Add cycle identifier to each dataset
cchs2001 <- cchs2001 %>% mutate(cycle = "2001")
cchs2003 <- cchs2003 %>% mutate(cycle = "2003")
cchs2005 <- cchs2005 %>% mutate(cycle = "2005")
cchs2007_2008 <- cchs2007_2008 %>% mutate(cycle = "2007-2008")
cchs2009_2010 <- cchs2009_2010 %>% mutate(cycle = "2009-2010")
cchs2011_2012 <- cchs2011_2012 %>% mutate(cycle = "2011-2012")
cchs2013_2014 <- cchs2013_2014 %>% mutate(cycle = "2013-2014")

# Create a list of all datasets
cchs_list <- list(
  cchs2001, cchs2003, cchs2005, cchs2007_2008, 
  cchs2009_2010, cchs2011_2012, cchs2013_2014
)
```

# Variable harmonization

## Core smoking variables

For the CSHM, we need to harmonize several key smoking-related variables:

| Type | Variables | Description |
|------|-----------|-------------|
| Status | SMK_01A, SMK_202, SMKDSTY | Basic smoking status indicators |
| Initiation | SMKG01C_cont, SMKG203_cont, SMKG207_cont | Age of smoking initiation |
| Cessation | SMK_09A_cont, SMKG09C | When smoking cessation occurred |
| Intensity | SMK_204, SMK_208, SMK_05B, SMK_05C | Quantity of cigarettes smoked |

## Step 4: Harmonize variables using cchsflow

The `cchsflow` package provides tools for harmonizing CCHS variables across cycles:

```{r harmonize, eval=FALSE}
# Define the variables needed for CSHM
smoking_vars <- c(
  "SMK_01A",        # In lifetime, smoked 100 or more cigarettes
  "SMK_202",        # Current smoking status
  "SMKDSTY",        # Type of smoker
  "SMKG01C_cont",   # Age smoked first cigarette
  "SMKG203_cont",   # Age started smoking daily (daily smokers)
  "SMKG207_cont",   # Age started smoking daily (former daily smokers)
  "SMK_09A_cont",   # When stopped smoking daily (former daily)
  "SMKG09C",        # Years since stopped smoking daily
  "SMK_204",        # Number of cigarettes smoked daily (daily smokers)
  "SMK_208",        # Number of cigarettes smoked daily (former daily smokers)
  "SMK_05B",        # Number of cigarettes smoked daily (occasional smokers)
  "SMK_05C"         # Number of days smoked in past month (occasional smokers)
)

# Demographic variables
demographic_vars <- c(
  "DHH_SEX",        # Sex of respondent
  "DHH_AGE",        # Age of respondent
  "WTS_M"           # Survey weight
)

# Variables for timing
timing_vars <- c(
  "SAMPLEID"        # Contains year and month of survey
)

# Combine all variables
all_vars <- c(smoking_vars, demographic_vars, timing_vars)

# Process each dataset with cchsflow
harmonized_list <- lapply(cchs_list, function(dataset) {
  rec_with_table(dataset, all_vars)
})

# Combine all harmonized datasets
harmonized_data <- bind_rows(harmonized_list)
```

## Step 5: Create derived variables

Some additional variables need to be created for the CSHM:

```{r derived-vars, eval=FALSE}
# Create survey date and additional variables
harmonized_data <- harmonized_data %>%
  mutate(
    # Extract survey year and month from SAMPLEID
    survey_year = as.numeric(substr(as.character(SAMPLEID), 1, 4)),
    survey_month = as.numeric(substr(as.character(SAMPLEID), 5, 6)),
    
    # Create a date for the survey (use middle of month as approximation)
    cchsbdate = as.Date(paste(survey_year, survey_month, "15", sep = "-")),
    
    # Rename age of first cigarette for consistency with original code
    agefirst = SMKG01C_cont,
    
    # Create survey age variable (age at time of survey)
    surveyage = DHH_AGE,
    
    # Calculate birth year
    birth_year = survey_year - surveyage,
    
    # Create weight variable with consistent name
    weighting = WTS_M,
    
    # Create respondent ID
    ont_id = row_number()  # Create sequential ID if none exists
  )
```

## Step 6: Clean and validate the data

Data cleaning is essential to ensure valid analyses:

```{r cleaning, eval=FALSE}
# Remove cases with missing key variables
cleaned_data <- harmonized_data %>%
  filter(!is.na(DHH_SEX)) %>%         # Remove cases missing sex
  filter(!is.na(surveyage)) %>%       # Remove cases missing age
  filter(surveyage >= 12) %>%         # Include only respondents aged 12+
  filter(!is.na(weighting))           # Remove cases missing weights

# Recode sex to match CSHM requirements (M/F)
cleaned_data <- cleaned_data %>%
  mutate(sex = ifelse(DHH_SEX == 1, "M", "F"))
```

## Step 7: Check variable distributions

Verify that the harmonization worked correctly by checking key variable distributions:

```{r check-distributions, eval=FALSE}
# Check smoking status distribution
cleaned_data %>%
  count(SMKDSTY) %>%
  mutate(
    label = case_when(
      SMKDSTY == 1 ~ "Daily smoker",
      SMKDSTY == 2 ~ "Occasional smoker",
      SMKDSTY == 3 ~ "Former daily smoker",
      SMKDSTY == 4 ~ "Former occasional smoker", 
      SMKDSTY == 5 ~ "Never smoker",
      TRUE ~ "Unknown"
    ),
    percentage = 100 * n / sum(n)
  ) %>%
  arrange(SMKDSTY)

# Check distribution by cycle
cleaned_data %>%
  group_by(cycle) %>%
  count() %>%
  mutate(percentage = 100 * n / sum(n))

# Check age distribution
cleaned_data %>%
  mutate(age_group = cut(surveyage, 
                       breaks = c(12, 20, 30, 40, 50, 60, 70, 80, 999),
                       labels = c("12-19", "20-29", "30-39", "40-49", 
                                 "50-59", "60-69", "70-79", "80+"))) %>%
  count(age_group) %>%
  mutate(percentage = 100 * n / sum(n))
```

# Save the processed data

Once your data is harmonized and validated, save it for future use:

```{r save-data, eval=FALSE}
# Save the processed data
saveRDS(cleaned_data, file.path(data_path, "harmonized_cchs_data.rds"))
```

# Additional considerations

## Variable mapping challenges

CCHS variables may not always harmonize cleanly. Here are some common challenges and solutions:

1. **Variable availability**: Not all variables are available in all cycles. Check which cycles include your variables of interest.

2. **Coding differences**: Response codes may differ across cycles, even after harmonization. Verify code mappings using `cchsflow` documentation.

3. **Skip patterns**: Survey skip patterns have changed over time, affecting the population asked certain questions. 

4. **Variable granularity**: Some cycles may have more detailed responses than others. Harmonization usually adopts the least granular version.

## Working with secure environments

When working in Statistics Canada Research Data Centres (RDCs) or other secure environments:

1. Use relative file paths or configuration files to manage data locations
2. Avoid storing large intermediate files if storage is limited
3. Follow the RDC's guidelines for output review
4. Consider batch processing if interactive sessions are limited

# Next steps

After harmonizing your CCHS data, you can proceed to:

1. [Process smoking variables](processing-cchs-data.qmd) in more detail
2. Begin [APC modeling](reproducing-manuel-study.qmd) using the harmonized data

# References

Canadian Community Health Survey (CCHS) - Annual Component. Statistics Canada.

Onysko, J., & MacKenzie, A. (2015). The cchsflow R package: Standardizing variables across CCHS cycles. Statistics Canada.