---
title: "Table 1 — Study population characteristics"
author: "CSHM Development Team"
date: today
format:
  html:
    toc: true
    self-contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))

cfg <- config::get(file = "config.yml")
```

# Table 1a — Study population before imputation

Characteristics of CCHS respondents included in CSHM, prior to multiple imputation.
Columns show males and females separately; overall column shows the combined sample.

```{r table-1a}
source("R/variables-sheet-utils.R")
source("R/variable-details-sheet-utils.R")
source("R/get-descriptive-data.R")
source("R/create-descriptive-tables.R")
source("R/descriptive-data.R")

variables_sheet        <- targets::tar_read("variables_sheet")
variable_details_sheet <- targets::tar_read("variable_details_sheet")
table_1a_data          <- targets::tar_read("table_1a_data")

# Table rows: the table1 role (includes auxiliary covariates that are not APC predictors)
table1_vars <- select_vars_by_role("table1", variables_sheet)

create_descriptive_table(
  descriptive_data      = table_1a_data,
  variables_sheet       = variables_sheet,
  variable_details_sheet = variable_details_sheet,
  variables             = table1_vars,
  column_stratifier     = survey_var(cfg, "sex"),
  sections_order        = c("Sociodemographics", "Health status", "Health behaviour")
)
```

# Table 1b — Study population after imputation

```{r table-1b}
table_1b_data <- targets::tar_read("table_1b_data")

create_descriptive_table(
  descriptive_data      = table_1b_data,
  variables_sheet       = variables_sheet,
  variable_details_sheet = variable_details_sheet,
  variables             = table1_vars,
  column_stratifier     = survey_var(cfg, "sex"),
  sections_order        = c("Sociodemographics", "Health status", "Health behaviour")
)
```

# Appendix — Characteristics by survey cycle (pre-imputation)

Each cycle shows sex-stratified results. Used to assess temporal trends in
sample composition and smoking variable availability across cycles.

```{r table-cycle}
cleaned_data <- targets::tar_read("cleaned_data")

create_cycle_specific_descriptive_table(
  study_data            = cleaned_data,
  variables_sheet       = variables_sheet,
  variable_details_sheet = variable_details_sheet,
  variables             = table1_vars,
  cycle_col             = survey_var(cfg, "cycle"),
  cycle_labels          = cfg$cycle_labels,
  column_stratifier     = survey_var(cfg, "sex"),
  weight_var            = survey_var(cfg, "weight"),
  sections_order        = c("Sociodemographics", "Health status", "Health behaviour")
)
```

# Appendix — Fully unweighted Tables 1a and 1b

Unweighted percentages and quantiles for comparison with the survey-weighted
presentation above (protocol §3.4.1).

```{r table-1-unweighted}
table_1a_unweighted <- targets::tar_read("table_1a_unweighted_data")
table_1b_unweighted <- targets::tar_read("table_1b_unweighted_data")

create_descriptive_table(
  descriptive_data      = table_1a_unweighted,
  variables_sheet       = variables_sheet,
  variable_details_sheet = variable_details_sheet,
  variables             = table1_vars,
  column_stratifier     = survey_var(cfg, "sex"),
  sections_order        = c("Sociodemographics", "Health status", "Health behaviour")
)

create_descriptive_table(
  descriptive_data      = table_1b_unweighted,
  variables_sheet       = variables_sheet,
  variable_details_sheet = variable_details_sheet,
  variables             = table1_vars,
  column_stratifier     = survey_var(cfg, "sex"),
  sections_order        = c("Sociodemographics", "Health status", "Health behaviour")
)
```
