---
title: "Reproducing the Canadian Smoking Histories Model"
subtitle: "A Guide to Variable Derivation and Analysis"
author: "CSHM Development Team"
date: "Last Updated: `r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    theme: cosmo
    code-fold: show
execute:
  echo: true
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Load required packages
library(dplyr)
library(ggplot2)
# If you have installed the cshgm package, uncomment this:
# library(cshm)
```

# Introduction

This document provides a guide to reproducing the Canadian Smoking Histories Model as described in @Manuel_HR_2020. The model uses Age-Period-Cohort (APC) analysis to estimate smoking initiation and cessation probabilities across different birth cohorts, which are then used to generate complete smoking histories.

## Purpose of This Guide

This guide aims to:

1.  Describe the key variables needed from the Canadian Community Health Survey (CCHS)
2.  Show how to derive the necessary variables for the model
3.  Provide R code implementations of the original SAS procedures
4.  Explain how to interpret and validate the results

## Original Study Overview

The original study [@Manuel_HR_2020] developed a model to:

-   Generate individual smoking histories for a simulated population
-   Estimate smoking prevalence across different time periods and birth cohorts
-   Model both smoking initiation and cessation probabilities
-   Account for age, period, and cohort effects in smoking behavior

The approach used data from the Canadian Community Health Survey to construct smoking histories and applied statistical modeling to estimate probabilities of starting and quitting smoking.

# Required Data Sources

## Canadian Community Health Survey (CCHS)

The CCHS is the primary data source for this analysis. Multiple cycles have been conducted:

-   CCHS 2001 (Cycle 1.1)
-   CCHS 2003 (Cycle 2.1)
-   CCHS 2005 (Cycle 3.1)
-   CCHS 2007-2008
-   CCHS 2009-2010
-   CCHS 2011-2012
-   CCHS 2013-2014
-   CCHS 2015-2016
-   CCHS 2017-2018

## Variable Harmonization Across Cycles

CCHS variable names have changed across cycles. We use the cchsflow package for variable harmonization:

```{r harmonization-example, eval=FALSE}
# Example of harmonizing CCHS data using cchsflow
# library(cchsflow)
# harmonized_data <- rec_with_table(
#   cchs_data,
#   c("SMK_01A", "SMKG01C_cont", "SMKG203_cont", "SMK_204") 
# )
```

# Key Variables for Smoking History

## Core Smoking Status Variables

The following variables are critical for determining smoking status:

| Harmonized Variable | Description | Values |
|----------------------------------|---------------------|-----------------|
| SMK_01A | In lifetime, smoked 100 or more cigarettes | 1=Yes, 2=No |
| SMK_202 | Current smoking status | 1=Daily, 2=Occasional, 3=Not at all |
| SMKDSTY | Type of smoker | 1=Daily, 2=Occasional, 3=Former daily, 4=Former occasional, 5=Never |

## Smoking Initiation Variables

Variables measuring when someone started smoking:

| Harmonized Variable | Description | Type |
|-----------------------------------|---------------------|----------------|
| SMKG01C_cont | Age smoked first cigarette | Continuous |
| SMKG203_cont | Age started smoking daily (daily smokers) | Continuous |
| SMKG207_cont | Age started smoking daily (former daily smokers) | Continuous |

## Smoking Cessation Variables

Variables measuring when someone quit smoking:

| Harmonized Variable | Description | Type |
|-----------------------------------|---------------------|----------------|
| SMK_09A_cont | When stopped smoking daily (former daily) | Continuous |
| SMKG09C | Years since stopped smoking daily (former daily) | Categorical |

## Smoking Intensity Variables

Variables measuring how much someone smokes/smoked:

| Harmonized Variable | Description | Type |
|-----------------------------------|---------------------|----------------|
| SMK_204 | Number of cigarettes smoked daily (daily smokers) | Continuous |
| SMK_208 | Number of cigarettes smoked daily (former daily smokers) | Continuous |
| SMK_05B | Number of cigarettes smoked daily (occasional smokers) | Continuous |
| SMK_05C | Number of days smoked in past month (occasional smokers) | Continuous |

## Demographic Variables

Variables used for cohort assignment and stratification:

| Harmonized Variable | Description       | Type        |
|---------------------|-------------------|-------------|
| DHH_SEX             | Sex of respondent | Categorical |
| cchsbdate           | CCHS survey date  | Date        |

# Variable Derivation for Age-Period-Cohort Analysis

## Smoking Initiation Process

### Step 1: Identify Smoking Initiators

The first step is to identify individuals who have initiated smoking, defined as having smoked at least 100 cigarettes in their lifetime.

```{r smoking-initiators, eval=FALSE}
# Using our process_smoking_initiation function
inits_male <- process_smoking_initiation(harmonized_data, sex = "M")
inits_female <- process_smoking_initiation(harmonized_data, sex = "F")
```

The function shown above performs the following steps:

1.  Filters respondents by sex
2.  Identifies never-smokers (SMK_01A = 2) and sets their initiation age to 101 (a flag value)
3.  Creates an initiation indicator (init=1 for smokers, init=0 for never-smokers)
4.  Filters to include only those born in 1920 or later
5.  Creates a subset containing only those who initiated smoking
6.  Applies age filter (age \>= 8)
7.  Calculates period (year of initiation) as cohort + age
8.  Returns a dataset with ont_id, weighting, age, cohort, period, and init variables

### Step 2: Create Population Denominators

To calculate initiation rates, we need the population at risk of initiation at each age, period, and cohort combination.

```{r denominator-example, eval=FALSE}
# Example function for creating denominators (not fully implemented yet)
create_initiation_denominators <- function(dataset, sex = "M") {
  # Function implementation would go here
  # This creates population counts by age, period, and cohort
}
```

## Smoking Cessation Process

### Step 1: Identify Smoking Cessation

Similarly, we identify individuals who have quit smoking and when they quit.

```{r cessation-example, eval=FALSE}
# Example function for processing cessation data
process_smoking_cessation <- function(dataset, sex = "M") {
  # Function implementation would follow similar steps to initiation
  # but with cessation-specific criteria
}
```

## Age-Period-Cohort Modeling

The APC modeling approach uses splines to model age, period, and cohort effects:

1.  Natural cubic splines are constructed for age, period, and cohort
2.  These splines are used in a generalized linear model with binomial errors
3.  Survival probabilities are used to adjust for differential mortality
4.  Separate models are fit for males and females

```{r apc-splines, eval=FALSE}
# Example of creating splines for APC modeling
create_age_splines <- function(ages, knots) {
  # Function to create natural cubic splines for ages
  # Implementation would go here
}
```

# Applying the Model

## Estimating Smoking Probabilities

Once the APC models are fitted, they can be used to predict:

1.  Initiation probabilities by age, period, and cohort
2.  Cessation probabilities by age, period, and cohort

```{r predict-example, eval=FALSE}
# Example of predicting initiation probabilities
predict_initiation <- function(age, period, cohort, sex) {
  # Implementation would use fitted coefficients from APC model
}
```

## Generating Smoking Histories

Complete smoking histories can be generated using:

1.  A microsimulation approach based on the estimated probabilities
2.  Monte Carlo methods to introduce appropriate stochasticity

```{r history-example, eval=FALSE}
# Example of generating a smoking history
generate_smoking_history <- function(birth_year, sex) {
  # Simulation implementation would go here
}
```

# Validation and Calibration

## Comparing to Historical Data

Model validation involves comparing the generated smoking histories to:

1.  Historical smoking prevalence data
2.  Known patterns in smoking behavior by birth cohort
3.  Other published estimates of smoking trends

```{r validation-example, eval=FALSE}
# Example of validating model results
validate_smoking_prevalence <- function(simulated_data, reference_data) {
  # Validation code would go here
}
```

## Calibration Procedures

If necessary, model calibration can be performed by:

1.  Adjusting the baseline rates
2.  Modifying the period effects
3.  Fine-tuning cohort-specific parameters

# Implementation Example

Below is a simplified example of processing smoking initiation data using the function we've developed:

```{r implementation-example, eval=FALSE}
# Create simulated data for testing
test_data <- data.frame(
  ont_id = 1001:1010,
  sex = rep(c("M", "F"), 5),
  SMK_01A = c(1, 2, 1, 1, 1, 2, 1, 2, 1, 1),  # 1=Yes, 2=No to 100+ cigarettes
  agefirst = c(16, NA, 12, 21, 7, NA, 18, NA, 15, 22),
  cchsbdate = as.Date(c("2001-06-15", "2001-07-20", "2001-08-10", 
                      "2001-09-05", "2001-10-25", "2002-01-15",
                      "2002-02-20", "2002-03-10", "2002-04-15", "2002-05-20")),
  weighting = round(runif(10, 100, 300))
)

# Process the initiation data
male_init <- process_smoking_initiation(test_data, sex = "M")
female_init <- process_smoking_initiation(test_data, sex = "F")

# Examine the results
print(male_init)
print(female_init)
```

# Conclusion

This guide has provided an overview of how to:

1.  Prepare CCHS data for smoking history analysis
2.  Derive the key variables needed for APC modeling
3.  Implement the core components of the Canadian Smoking Histories Model
4.  Apply and validate the model

Further development is ongoing to complete the full implementation of the model in R, including additional validation against historical data and extensions to incorporate more recent survey cycles.

# References

::: {#refs}
:::

# Appendix: Additional Resources

## SAS to R Translation Notes

The original model was implemented in SAS. Key translation considerations include:

1.  SAS `PROC GENMOD` is equivalent to R's `glm()` function
2.  SAS natural cubic splines can be created with R's `splines::ns()`
3.  SAS data steps are translated to R's data.frame operations
4.  SAS `%macro` functions are converted to R functions

## Complete Variable Crosswalk

For a complete mapping of variables across CCHS cycles, refer to the included variable sheets:

-   `worksheets/cshm-variables.csv`
-   `worksheets/cshm-variable-details.csv`