---
title: "Stage 1 — Variable setup"
subtitle: "Study data dictionary and variable roles"
---

## Purpose

This page serves as the **study data dictionary** — a complete description of every variable used in the CSHM, generated directly from those cchsflow worksheets. If the cchsflow worksheets change, this page updates automatically on re-render. 

The data includes both PUMF and Master versions of variables available for all CCHS cycles. Variables in cchsflow are harmonized across cycles. There are many harmonization issues that are discussed in the cchsflow R package, as well as the protocol. Further the CCHS survey major undergoes redevelopment at different periods. Redevelopment includes redevelopment of the study frame and selection process that affects response bias.

The main different between the PUMF and Main versions of the CCHS is categorizing continuous variables like age to reduce the risk of responder identification. In this study, the categorized groups like age (DHHGAGE identified as a categorical variable by the "G" before "AGE") are transformed into "pseudo-continous variables" (DHHGAGE_cont identified by "_cont") by assiging the mid-point value of the cateogy. 

## Inputs

| File | Contents |
|------|----------|
| `worksheets/cshm-variables.csv` | Study variable list: name, label, type, role, section, source |
| `worksheets/cchsflow-variable-details.csv` | In-repo snapshot of cchsflow v3 base recoding rules (cycles 2001–2023) |
| `worksheets/cshm-variable-details.csv` | CSHM extension rows (GEOGPRV and WTS_M, 2019–20 and 2022 cycles) |

## Outputs

| Target | Description |
|--------|-------------|
| `variables_sheet` | Data frame: one row per study variable |
| `variable_details_sheet` | Data frame: combined cchsflow base + CSHM extension recoding rules |

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# root must be set before source() calls; use absolute paths here
repo_root <- normalizePath("../..")
knitr::opts_knit$set(root.dir = repo_root)

cfg    <- config::get(file = file.path(repo_root, "config.yml"))
source(file.path(repo_root, "R/variables-sheet-utils.R"))
source(file.path(repo_root, "R/variable-details-sheet-utils.R"))

withr::with_dir(repo_root, {
  variables_sheet        <- targets::tar_read("variables_sheet")
  variable_details_sheet <- targets::tar_read("variable_details_sheet")
})
```

```{r helpers, include=FALSE}
library(dplyr)
library(knitr)

# All cycles used in this study (PUMF)
all_pumf_cycles <- c("2001", "2003", "2005", "2007-08", "2009-10",
                     "2011-12", "2013-14", "2015-16", "2017-18", "2019-20", "2022")

pumf_cycle_map <- c(
  cchs2001_p = "2001", cchs2003_p = "2003", cchs2005_p = "2005",
  cchs2007_2008_p = "2007-08", cchs2009_2010_p = "2009-10",
  cchs2011_2012_p = "2011-12", cchs2013_2014_p = "2013-14",
  cchs2015_2016_p = "2015-16", cchs2017_2018_p = "2017-18",
  cchs2019_2020_p = "2019-20", cchs2022_p = "2022"
)
# Master coverage: cchsflow v3 uses _m suffix for Master/RDC files.
# Legacy _i (ICES) and _s (share) suffixes were removed in v3.
master_cycle_map <- c(
  cchs2001_m = "2001", cchs2003_m = "2003", cchs2005_m = "2005",
  cchs2007_2008_m = "2007-08", cchs2009_2010_m = "2009-10",
  cchs2011_2012_m = "2011-12", cchs2013_2014_m = "2013-14",
  cchs2015_2016_m = "2015-16", cchs2017_2018_m = "2017-18",
  cchs2019_2020_m = "2019-20", cchs2021_m = "2021",
  cchs2022_m = "2022", cchs2023_m = "2023"
)

# --- helper: extract variable names from variableStart notation ---
# Returns list(pumf = "VARNAME", master = "MASTERNAME or same")
extract_var_names <- function(var_name, variable_start) {
  if (is.na(variable_start) || variable_start == "") return(list(pumf = var_name, master = ""))
  # fallback name in [brackets] is the canonical (master-file) name
  fallback <- regmatches(variable_start, regexpr("\\[([A-Za-z0-9_]+)\\]", variable_start))
  fallback <- if (length(fallback) > 0) gsub("\\[|\\]", "", fallback) else var_name
  # if fallback differs from var_name, it's the master variable name
  master_name <- if (fallback != var_name) fallback else ""
  list(pumf = var_name, master = master_name)
}

# --- helper: get cycle coverage from variable_details ---
# Returns list(pumf = "all" or "2001, 2003 ...", master = "all" or "...")
get_cycle_coverage <- function(var_name, variable_details_sheet) {
  rows <- variable_details_sheet[variable_details_sheet$variable == var_name, ]
  if (nrow(rows) == 0) return(list(pumf = "—", master = "—"))

  # expand databaseStart (comma-separated dataset names) across all rows
  all_datasets <- unlist(strsplit(as.character(rows$databaseStart), ",\\s*"))
  all_datasets <- trimws(all_datasets)

  pumf_yrs   <- unique(pumf_cycle_map[all_datasets[all_datasets %in% names(pumf_cycle_map)]])
  master_yrs <- unique(master_cycle_map[all_datasets[all_datasets %in% names(master_cycle_map)]])
  pumf_yrs   <- sort(pumf_yrs[!is.na(pumf_yrs)])
  master_yrs <- sort(master_yrs[!is.na(master_yrs)])

  pumf_str <- if (length(pumf_yrs) == 0) "—"
              else if (identical(pumf_yrs, all_pumf_cycles)) "all 11 cycles"
              else paste(pumf_yrs, collapse = ", ")

  master_str <- if (length(master_yrs) == 0) "—"
                else paste(master_yrs, collapse = ", ")

  list(pumf = pumf_str, master = master_str)
}

# --- helper: categories as HTML lines (one per line) ---
get_categories_html <- function(var_name, variable_details_sheet) {
  rows <- get_unique_rec_end_rows(variable_details_sheet, var_name, include_NA = FALSE)
  if (nrow(rows) == 0) return("")
  pairs <- ifelse(
    is.na(rows$catLabel) | rows$catLabel == "",
    as.character(rows$recEnd),
    paste0("<b>", rows$recEnd, "</b>: ", rows$catLabel)
  )
  paste(pairs, collapse = "<br>")
}

# --- helper: continuous range ---
get_range <- function(var_name, variable_details_sheet) {
  rows <- variable_details_sheet[variable_details_sheet$variable == var_name, ]
  rows <- rows[!grepl("^NA::", as.character(rows$recEnd)), ]
  if (nrow(rows) == 0) return("")
  ends <- suppressWarnings(as.numeric(as.character(rows$recEnd)))
  ends <- ends[!is.na(ends)]
  if (length(ends) == 0) return("")
  paste0(min(ends), "–", max(ends))
}

# --- helper: missing data codes as HTML lines ---
get_na_codes_html <- function(var_name, variable_details_sheet) {
  rows <- variable_details_sheet[variable_details_sheet$variable == var_name, ]
  na_rows <- rows[grepl("^NA::", as.character(rows$recEnd)), ]
  if (nrow(na_rows) == 0) return("—")
  codes <- paste0("<b>", na_rows$recEnd, "</b>: ", na_rows$catLabel)
  paste(unique(codes), collapse = "<br>")
}

# --- helper: cycle coverage as named vectors for matrix table ---
# Returns list(pumf = named logical, master = named logical) over all_cycles.
# master_var_name: when the Master file uses a different variable name (e.g.,
# DHH_AGE for PUMF's DHHGAGE_cont), pass it here to look up Master coverage
# from that variable's rows in variable_details_sheet.
all_cycles <- c("2001", "2003", "2005", "2007-08", "2009-10",
                "2011-12", "2013-14", "2015-16", "2017-18", "2019-20", "2022", "2023")

get_cycle_coverage_vec <- function(var_name, variable_details_sheet,
                                   master_var_name = NULL) {
  # PUMF coverage: always from the primary variable name
  rows <- variable_details_sheet[variable_details_sheet$variable == var_name, ]
  if (nrow(rows) == 0) {
    pumf_v <- setNames(rep(FALSE, length(all_cycles)), all_cycles)
  } else {
    all_datasets <- trimws(unlist(strsplit(as.character(rows$databaseStart), ",\\s*")))
    pumf_yrs <- unname(pumf_cycle_map[all_datasets[all_datasets %in% names(pumf_cycle_map)]])
    pumf_v   <- setNames(all_cycles %in% pumf_yrs, all_cycles)
  }

  # Master coverage: use master_var_name if provided, otherwise same variable
  master_lookup <- if (!is.null(master_var_name) && master_var_name != var_name) {
    master_var_name
  } else {
    var_name
  }
  m_rows <- variable_details_sheet[variable_details_sheet$variable == master_lookup, ]
  if (nrow(m_rows) == 0) {
    # Also check primary variable rows for _m datasets (variables with same name in both)
    m_rows <- rows
  }
  if (nrow(m_rows) == 0) {
    master_v <- setNames(rep(FALSE, length(all_cycles)), all_cycles)
  } else {
    m_datasets <- trimws(unlist(strsplit(as.character(m_rows$databaseStart), ",\\s*")))
    master_yrs <- unname(master_cycle_map[m_datasets[m_datasets %in% names(master_cycle_map)]])
    master_v   <- setNames(all_cycles %in% master_yrs, all_cycles)
  }

  list(pumf = pumf_v, master = master_v)
}

# --- build config key → pumf/master variable mapping from cfg$survey ---
# Scalar entries (e.g. cycle) are skipped; only pumf/master list entries are kept.
# Each source entry is now a list with $var, $min, $max.
`%||%` <- function(x, y) if (is.null(x)) y else x
survey_cfg <- cfg$survey
survey_map <- do.call(rbind, lapply(names(survey_cfg), function(key) {
  entry <- survey_cfg[[key]]
  if (!is.list(entry)) return(NULL)  # skip scalars like cycle
  pumf   <- entry[["pumf"]]
  master <- entry[["master"]]
  data.frame(
    config_key     = key,
    pumf_varname   = if (is.list(pumf))   pumf[["var"]]   else pumf   %||% NA_character_,
    master_varname = if (is.list(master)) master[["var"]] else master %||% NA_character_,
    stringsAsFactors = FALSE
  )
}))
```

## Study data dictionary

Each row is one variable. The **Config key : CCHS variable** column shows the `config.yml` key paired with the PUMF variable name; where the Master file uses a different name it is shown on a second line. Categories and ranges are from `variable_details_sheet`.

^[Missing data codes apply uniformly to all variables: **NA(a)** = not applicable; **NA(b)** = don't know or refused; **NA(c)** = not asked in this cycle.]

```{r data-dictionary}
#| results: asis
design_vars <- variables_sheet$variable[grepl("design", variables_sheet$role)]

dict <- variables_sheet |>
  filter(!variable %in% design_vars) |>
  rowwise() |>
  mutate(
    config_key  = {
      hit <- survey_map$config_key[!is.na(survey_map$pumf_varname) &
                                     survey_map$pumf_varname == variable]
      if (length(hit) > 0) hit[1] else ""
    },
    master_name = {
      hit <- survey_map$master_varname[!is.na(survey_map$pumf_varname) &
                                         survey_map$pumf_varname == variable]
      if (length(hit) > 0 && !is.na(hit[1]) && hit[1] != variable) hit[1] else ""
    },
    key_var     = {
      key <- config_key
      pumf_part <- if (key != "") paste0(key, ": ", variable) else variable
      if (master_name != "" && master_name != variable)
        paste0(pumf_part, "<br><small>Master: ", master_name, "</small>")
      else
        pumf_part
    },
    values      = if (variableType == "Categorical")
                    get_categories_html(variable, variable_details_sheet)
                  else
                    get_range(variable, variable_details_sheet)
  ) |>
  ungroup() |>
  select(section, key_var, label, variableType, units, values)

sections <- unique(dict$section)
sections <- sections[!is.na(sections)]

for (sec in sections) {
  cat("\n\n###", sec, "\n\n")
  tbl <- dict |>
    filter(section == sec) |>
    select(-section) |>
    rename(
      `Config key: CCHS variable` = key_var,
      Label            = label,
      Type             = variableType,
      Units            = units,
      `Values / range` = values
    )
  print(kable(tbl, format = "html", escape = FALSE,
              table.attr = 'class="table table-sm table-striped"'))
}
```

## Cycle coverage

Each cell shows whether the variable is available in that survey cycle. **P** = PUMF, **M** = Master, **P+M** = both. Blank = not available. Gaps indicate cycles where the variable was not asked or not yet harmonized.

```{r cycle-coverage}
#| results: asis
# design_vars already defined in data-dictionary chunk above
cov_rows <- variables_sheet |>
  filter(!variable %in% design_vars) |>
  rowwise() |>
  mutate(
    config_key = {
      hit <- survey_map$config_key[!is.na(survey_map$pumf_varname) &
                                     survey_map$pumf_varname == variable]
      if (length(hit) > 0) hit[1] else ""
    },
    master_var = {
      hit <- survey_map$master_varname[!is.na(survey_map$pumf_varname) &
                                         survey_map$pumf_varname == variable]
      if (length(hit) > 0 && !is.na(hit[1]) && hit[1] != "") hit[1] else NA_character_
    },
    cov = list(get_cycle_coverage_vec(
      variable, variable_details_sheet,
      master_var_name = if (is.na(master_var)) NULL else master_var
    ))
  ) |>
  ungroup()

sections <- unique(cov_rows$section)
sections <- sections[!is.na(sections)]

for (sec in sections) {
  cat("\n\n###", sec, "\n\n")
  sec_rows <- cov_rows[!is.na(cov_rows$section) & cov_rows$section == sec, ]

  mat <- do.call(rbind, lapply(seq_len(nrow(sec_rows)), function(i) {
    cv <- sec_rows$cov[[i]]
    sapply(all_cycles, function(yr) {
      p <- cv$pumf[yr]
      m <- cv$master[yr]
      if (isTRUE(p) && isTRUE(m)) "P+M"
      else if (isTRUE(p))          "P"
      else if (isTRUE(m))          "M"
      else                          ""
    })
  }))

  tbl <- data.frame(
    `Config key` = sec_rows$config_key,
    Label        = sec_rows$label,
    mat,
    check.names  = FALSE,
    stringsAsFactors = FALSE
  )
  colnames(tbl)[1:2] <- c("Config key", "Label")

  print(kable(tbl, format = "html", escape = FALSE,
              table.attr = 'class="table table-sm table-striped"'))
}
```

## Variable roles

Each variable carries one or more **roles** — a comma-separated label in `variables_sheet` that tells the pipeline how to use that variable. Roles allow downstream stages to select exactly the variables they need without any hard-coding.

```{r role-table}
# Role definitions
role_defs <- data.frame(
  Role = c("design", "intermediate", "predictor", "model-stratifier",
           "table1", "table1-stratifier",
           "apc-numerator", "apc-denominator",
           "imputation-predictor", "sensitivity-analysis"),
  Group = c("Survey design", "Harmonization", "Model", "Model",
            "Descriptive", "Descriptive",
            "APC data prep", "APC data prep",
            "Imputation", "Analysis"),
  Purpose = c(
    "Survey infrastructure — cycle identifier, sampling weights",
    "Raw cchsflow input used to derive a unified variable; not used directly by pipeline code",
    "Covariate in the APC model or descriptive analysis",
    "Stratifies APC models into separate fits (e.g., sex)",
    "Appears as a row in Table 1 descriptive statistics",
    "Stratifies Table 1 columns (e.g., sex columns)",
    "Defines the event indicator for APC numerator cells (Stage 7)",
    "Constructs the at-risk person-year denominator (Stage 7)",
    "Included in the MICE imputation model (Stage 5)",
    "Used in sensitivity analyses only"
  ),
  stringsAsFactors = FALSE
)
kable(role_defs, format = "html",
      table.attr = 'class="table table-sm table-striped"')
```

```{r role-variable-crosswalk}
#| results: asis
# Which variables carry which roles?
cat("\n\n### Variables by role\n\n")

# Expand multi-role rows
role_rows <- lapply(seq_len(nrow(variables_sheet)), function(i) {
  roles <- trimws(strsplit(variables_sheet$role[i], ",")[[1]])
  data.frame(role     = roles,
             variable = variables_sheet$variable[i],
             label    = variables_sheet$label[i],
             source   = variables_sheet$source[i],
             stringsAsFactors = FALSE)
})
role_expanded <- do.call(rbind, role_rows)

for (r in sort(unique(role_expanded$role))) {
  cat("\n\n**`", r, "`**\n\n", sep = "")
  vars <- role_expanded[role_expanded$role == r, c("variable", "label", "source")]
  print(kable(vars, format = "html", row.names = FALSE,
              table.attr = 'class="table table-sm"'))
}
```

## Key decisions

**Separation of definition from code.** Variable names, types, categories, and roles are all stored in plain-text CSVs. R code never contains a bare CCHS variable name — it always reads from `survey_var(cfg, "key")` (for key variables) or filters `variables_sheet` by role. This makes the pipeline adaptable to other surveys.

**PUMF vs Master.** The `source` column (`pumf` / `master` / `both`) controls which variables are loaded in each computing environment. `load_study_data()` filters at load time — PUMF variables only in the open environment; Master variables at the Statistics Canada RDC.

**Extension rows.** cchsflow v3 now covers age and sex (`DHH_SEX`, `DHHGAGE_cont`) for all 2001–2023 cycles, so their former extension rows have been removed. The CSHM extension rows (`worksheets/cshm-variable-details.csv`) still add recoding rules for `GEOGPRV` and `WTS_M` for the 2019–20 and 2022 cycles, which cchsflow v3 does not yet harmonize. As cchsflow extends its coverage, the remaining extension rows can be removed.

**Missing data codes.** The CSHM uses `haven::tagged_na()` throughout, preserving the reason for missingness: `NA(a)` = not applicable, `NA(b)` = don't know or refused, `NA(c)` = not asked in this cycle. These are carried through all pipeline stages and used in imputation (Stage 5).

## Related resources

- [Stage 2 — Data loading](2-data-loading.qmd)
- [Variable reference](../reference/variables.qmd)
- [How to adapt variables for a different survey](../how-to/custom-models.qmd)
- `schemas/cshm-variables.yaml` — role vocabulary (LinkML schema)
