---
title: "Stage 5 — Multiple imputation"
subtitle: "Handling missing data with where-matrix MICE"
---

## Purpose

The CCHS has missingness from two distinct mechanisms: item non-response (a respondent was asked but did not answer — tagged `NA(b)`) and structural missingness (the question did not apply — `NA(a)` — or was not asked in that cycle — `NA(c)`). Complete-case analysis would discard respondents and introduce selection bias; imputing *structural* missingness would invent values that should not exist (for example, a quit age for a never-smoker).

This stage applies multiple imputation by chained equations (MICE) to **item non-response only**. An explicit `where` matrix marks the `NA(b)` cells as the only cells MICE may fill; `NA(a)` and `NA(c)` survive the round trip untouched. The result is `m` completed datasets. The full specification, including the diagnostic and sensitivity-analysis plan, is the protocol's [Appendix D](../protocol/appendix-imputation.qmd).

## Inputs

| Target | Description |
|--------|-------------|
| `cleaned_data` | Age-restricted, truncated data frame (Stage 3) |
| `variables_sheet` | Variable metadata — `imputation-predictor` role identifies variables in the MICE model |

Configuration parameters:

| Parameter | Value | Meaning |
|-----------|-------|---------|
| `imputation_m` | 5 | Number of imputed datasets (1 in `dev`/`draft`) |
| `imputation_maxit` | 5 | MICE iterations per imputation |

## Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
cfg <- config::get(file = "config.yml")
```

```{r load-analysis-data}
# analysis_data is a list: datasets (m completed data frames), m,
# imputed_cells (audit of filled cells), logged_events (mice diagnostics)
analysis_data <- targets::tar_read("analysis_data")
```

```{r inspect-imputation}
# Which cells were imputed (NA(b) only, by construction)
analysis_data$imputed_cells

# Missingness before imputation, by type
cleaned_data    <- targets::tar_read("cleaned_data")
variables_sheet <- targets::tar_read("variables_sheet")

source("R/variables-sheet-utils.R")
imputation_vars <- select_vars_by_role("imputation-predictor", variables_sheet)
imputation_vars <- intersect(imputation_vars, colnames(cleaned_data))

# Proportion missing per variable (all NA types combined)
sort(sapply(cleaned_data[, imputation_vars], function(x) mean(is.na(x))), decreasing = TRUE)
```

## Outputs

| Target | Description |
|--------|-------------|
| `analysis_data` | List: `datasets` (the `m` completed data frames), `m`, `imputed_cells`, `logged_events` |

The completed datasets differ from `cleaned_data` only at the `NA(b)` cells. Table 1b (Stage 6) averages descriptive statistics across all `m` datasets; the APC stage (Stages 7–8) is fitted on imputation 1, with a documented upgrade path to per-imputation fits pooled by Rubin's rules (Appendix D).

## Key decisions

**`where`-matrix MICE.** `haven::tagged_na()` values are plain `NA` to `mice()`, so without the `where` matrix MICE would impute structural missingness. Only `NA(b)` cells are flagged for imputation; tagged `NA(a)`/`NA(c)` cells and the original factor levels are preserved by construction, because the write-back copies only `where`-matrix cells into the original data. If MICE declines to impute a flagged cell (constant or collinear variable), the stage stops with an error rather than silently degrading an `NA(b)` tag to plain `NA`.

**Design variables as predictors.** The survey cycle and sampling weight (`WTS_M`) enter the imputation model as predictors alongside age, sex, province, and education, making the imputation design-consistent and congenial with the APC analysis models.

**Imputation predictor selection.** Variables with the `imputation-predictor` role enter the MICE model. Derived variables (e.g., `pack_years_der`) are not imputed directly — they are recomputed from their imputed feeders.

**`m = 1` in dev/draft profiles.** The `dev` and `draft` config profiles use `imputation_m: 1` and `imputation_maxit: 1` for fast iteration. Production uses `m = 5, maxit = 5`; adequacy is checked against the fraction of missing information (Appendix D, Phase 2).

**Structural missingness is an analysis decision, not an imputation default.** `NA(c)` cycle-level absence (e.g., `age_start_smoking` in 2019–20 PUMF) is *not* imputed in the base pipeline. Whether to impute across cycle-level absence — a stronger MAR assumption — is evaluated with the delta-adjustment sensitivity analysis in Appendix D, with the RDC Master files as an external check.

## Related resources

- `R/imputation.R` — `impute_data()`, `prepare_for_mice()`
- [Protocol Appendix D — Missing data and imputation plan](../protocol/appendix-imputation.qmd)
- [mice R package](https://www.rdocumentation.org/packages/mice)
