---
title: "Stage 5 — Multiple imputation"
subtitle: "Handling missing data with MICE"
---

## Purpose

The CCHS has substantial missingness in sociodemographic variables, primarily because some variables are only collected in certain cycles (e.g., immigration status is absent from 2019–20 and 2022) or were refused/unknown by some respondents. Complete-case analysis would discard a large fraction of respondents and introduce selection bias.

Multiple imputation by chained equations (MICE) fills in missing values by iteratively fitting a model for each incomplete variable given all other variables. The result is `m = 5` complete datasets, which are later combined using Rubin's rules. This stage produces `analysis_data` — the primary input for all downstream modelling.

## Inputs

| Target | Description |
|--------|-------------|
| `cleaned_data` | Age-restricted, truncated data frame (Stage 3) |
| `variables_sheet` | Variable metadata — `imputation-predictor` role identifies variables for the MICE model |

Configuration parameters:

| Parameter | Value | Meaning |
|-----------|-------|---------|
| `imputation_m` | 5 | Number of imputed datasets |
| `imputation_maxit` | 5 | MICE iterations per imputation |

## Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
cfg <- config::get(file = "config.yml")
```

```{r load-analysis-data}
# analysis_data is a mids object (multiple imputation dataset)
analysis_data <- targets::tar_read("analysis_data")
```

```{r inspect-imputation}
# Summary of imputation
print(analysis_data)

# Missingness before imputation
cleaned_data    <- targets::tar_read("cleaned_data")
variables_sheet <- targets::tar_read("variables_sheet")

source("R/variables-sheet-utils.R")
imputation_vars <- select_vars_by_role("imputation-predictor", variables_sheet)
imputation_vars <- intersect(imputation_vars, colnames(cleaned_data))

# Proportion missing per variable
sort(sapply(cleaned_data[, imputation_vars], function(x) mean(is.na(x))), decreasing = TRUE)
```

## Outputs

| Target | Description |
|--------|-------------|
| `analysis_data` | `mids` object (mice): `m = 5` imputed datasets, stored as `data/analysis_data.rds` |

`analysis_data.rds` is the **primary reproducibility artefact** — sharing it allows others to reproduce all downstream analyses without access to the raw CCHS files.

## Key decisions

**MICE over single imputation.** Single imputation (e.g., mean substitution) underestimates variability and produces overconfident standard errors. MICE preserves uncertainty about missing values by propagating it across `m` imputed datasets and combining results with Rubin's rules.

**Imputation predictor selection.** Only variables with `role` containing `"imputation-predictor"` enter the MICE model. This excludes variables derived from others (e.g., `smoking_status` is derived from raw smoking items) and APC-specific construction variables, which would create circular dependencies.

**`m = 1` in dev/draft profiles.** The `dev` and `draft` config profiles use `imputation_m: 1` and `imputation_maxit: 1` to speed up iteration. The production run uses `m = 5, maxit = 5` per the protocol.

**Cycle-specific missingness.** Some variables (e.g., immigration status, ethnicity) are structurally absent from certain CCHS cycles. These are `NA(c)` (tagged NA: not asked this cycle) in the harmonized data. MICE treats these as missing and imputes them — this is appropriate because the variable *would* have had a value if collected. Tagged NA codes are stripped to regular `NA` before MICE is run.

## Related resources

- `R/imputation.R` — `impute_data()`
- [mice R package](https://www.rdocumentation.org/packages/mice)
- Protocol §3.3 — missing data strategy
