---
title: "Stage 6 — Descriptive statistics (post-imputation)"
subtitle: "Table 1b: study population characteristics after imputation"
---

## Purpose

After imputation, we repeat the descriptive statistics on the completed dataset. Comparing Table 1b to Table 1a (Stage 4) allows readers to see what imputation changed: proportions for variables with high missingness will shift, while variables with complete data should be unchanged.

This is standard practice in studies using multiple imputation — reporting both tables provides full transparency about the role of imputation in the analysis.

## Inputs

| Target | Description |
|--------|-------------|
| `analysis_data` | `mids` object — 5 imputed datasets (Stage 5) |
| `variables_sheet` | Variable metadata |
| `variable_details_sheet` | Recoding rules and value labels |

## Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
cfg <- config::get(file = "config.yml")
```

```{r load-table-data}
table_1b_data          <- targets::tar_read("table_1b_data")
variables_sheet        <- targets::tar_read("variables_sheet")
variable_details_sheet <- targets::tar_read("variable_details_sheet")
```

```{r render-table-1b}
source("R/variables-sheet-utils.R")
source("R/variable-details-sheet-utils.R")
source("R/get-descriptive-data.R")
source("R/create-descriptive-tables.R")

predictor_vars <- select_vars_by_role("predictor", variables_sheet)

create_descriptive_table(
  descriptive_data       = table_1b_data,
  variables_sheet        = variables_sheet,
  variable_details_sheet = variable_details_sheet,
  variables              = predictor_vars,
  column_stratifier      = survey_var(cfg, "sex"),
  sections_order         = c("Sociodemographics", "Health behaviour")
)
```

## Outputs

| Target | Description |
|--------|-------------|
| `table_1b_data` | Named list of post-imputation descriptive statistics |

## Key decisions

**Pooled over imputations.** `get_cshm_desc_data()` pools descriptive statistics across the `m = 5` imputed datasets using Rubin's rules for means and proportions. Standard deviations reflect within-imputation variability only (between-imputation variability in descriptives is negligible for reporting purposes).

**Same structure as Table 1a.** The same `create_descriptive_table()` call is used for both tables, ensuring the rows and columns are identical and differences are directly comparable.

## Related resources

- [Stage 4 — Pre-imputation descriptives](4-descriptive-statistics.qmd)
- `docs/results/table-1.qmd` — rendered Tables 1a and 1b side by side
