---
title: "Stage 7 — APC data preparation"
subtitle: "Building numerator and denominator cells for the Age-Period-Cohort models"
---

## Purpose

The APC models are logistic regressions where each row represents a *person-year at risk*. Before fitting the models, we need to convert the cross-sectional CCHS data into this person-year format by reconstructing each respondent's smoking history up to their survey date, then computing the denominator (person-years at risk of initiating or quitting) and numerator (events: initiations and cessations).

This is the most computationally expensive step. By keeping it as a separate target from Stage 8, `{targets}` can cache the APC dataset independently — changing model parameters (knots, constraints) only reruns Stage 8, not Stage 7.

## Inputs

| Target | Description |
|--------|-------------|
| `analysis_data` | Imputed dataset (Stage 5) |

Configuration used:

| Parameter | Location | Description |
|-----------|----------|-------------|
| `survey$age` | `config.yml` | Continuous age variable |
| `survey$sex` | `config.yml` | Sex variable |
| `survey$cycle` | `config.yml` | Survey cycle integer |
| `survey$weight` | `config.yml` | Sampling weight |
| `survey$smoking_status` | `config.yml` | 6-category smoking status |
| `survey$age_first_cigarette` | `config.yml` | Age at first whole cigarette |
| `survey$years_since_quit_complete` | `config.yml` | Years since stopping smoking completely (the cessation exit) |
| `survey$established_smoker` | `config.yml` | 100-cigarette criterion; `yes_code` is the value that includes a person in the smoking models |
| `survey$smoking_status` `*_codes` | `config.yml` | Which `SMKDSTY_original` codes count as ever, current, former, never |
| `survey$sex` `*_code` | `config.yml` | Sex codes used to stratify the models |
| `survey_bound(cfg, "age_first_cigarette", "min")` | `config.yml` | APC floor for initiation age (PUMF: 13, Master: 8) |
| `apc$cessation_durability_years` | `config.yml` | A quit counts as cessation only after this many years (2) |
| `apc$age_knots` | `config.yml` | `[10, 15, 20, 50, 60]` |
| `apc$period_knots` | `config.yml` | `[1940, 1950, 1960, 1970, 1980]` |
| `apc$cohort_knots` | `config.yml` | `[1930, 1940, 1945, 1950, 1955, 1960, 1965, 1970, 1975, 1980]` |

## Code

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../.."))
cfg <- config::get(file = "config.yml")
```

```{r load-apc-data}
apc_data <- targets::tar_read("apc_data")
# apc_data is a list: $initiation_men, $initiation_women,
#                      $cessation_men, $cessation_women
names(apc_data)
```

```{r inspect-apc-data}
# Initiation data: one row per person-year at risk of initiating
nrow(apc_data$initiation_men)
nrow(apc_data$initiation_women)

# Cessation data: one row per person-year at risk of quitting
nrow(apc_data$cessation_men)
nrow(apc_data$cessation_women)

# Event rates
mean(apc_data$initiation_men$init)
mean(apc_data$cessation_men$init)
```

## Outputs

| Target | Description |
|--------|-------------|
| `apc_data` | Named list with four data frames: `initiation_men`, `initiation_women`, `cessation_men`, `cessation_women` |

Each data frame contains:

| Column | Description |
|--------|-------------|
| `age` | Age at risk (years) |
| `period` | Calendar year |
| `cohort` | Birth year (`period - age`) |
| `event` | Event indicator (1 = initiated/quit, 0 = at risk but no event) |
| `weight` | Survey weight; multiplied by the mortality correction once one is implemented (currently none) |

The spline basis columns are built in Stage 8 (`build_spline_basis()`), not stored here. Each data frame carries two attributes set by `apply_survival_correction()`: `mortality_correction` (currently `"none"`) and `estimand_note`.

## Key decisions

**APC identity: `cohort = period − age`.** This fundamental identity is the basis for the APC framework. Each respondent's smoking history is reconstructed year-by-year using their reported age at initiation/cessation and the survey year. The cohort effect is not estimated independently — it is derived from the period and age effects.

**Retrospective history construction.** A current smoker at age 45 in survey year 2014 is recorded as a person-year at risk of cessation at every age from their initiation age to 45. This reconstruction assumes respondents accurately recall their age of initiation and cessation (supported by CCHS validation studies).

**Mortality adjustment.** Ever-smokers are less likely to survive to survey date than never-smokers, creating survival bias. No correction is applied yet: `cfg$apc$mortality_method` is `"none"`, the `weighting` column holds the survey weight alone, and the datasets carry an `estimand_note` attribute stating that results describe respondents who survived to be surveyed. MPoRT (primary) and Peto (sensitivity) are not yet implemented (protocol section 3.4.5).

**Cessation model: who is included and when time at risk begins.** The cessation model includes established smokers: people who have smoked 100 or more cigarettes in their lifetime (`smoked_100_lifetime`), whatever their current smoking pattern. The event is stopping smoking completely, dated by `time_quit_smoking_complete`. Each person's time at risk begins at their own age at first whole cigarette; the study floor age (`survey_bound(cfg, "age_first_cigarette", "min")`) is a reporting boundary only. A quit that has lasted fewer than `cfg$apc$cessation_durability_years` (2) years at the survey does not count as cessation: the person is a current smoker at the survey and their time at risk ends at the quit age without an event. A person who started and stopped at the same age contributes one year at risk with the event in it. People with a missing entry age or missing quit timing (including the whole 2001 cycle, where the stopped-completely questions were not asked) are excluded here and counted in the `cessation_diagnostics` attribute; imputation (task 1.8c) will supply their values. See `docs/development/estimand-specification.md`.

**Four separate data frames, not one.** Men and women are modelled separately (consistent with Manuel et al. 2020). Initiation and cessation are separate models. Keeping four data frames lets Stage 8 fit all four models in parallel using `{targets}` branching.

## Related resources

- [Stage 8 — APC model fitting](8-apc-model.qmd)
- [APC methodology explanation](../explanation/apc-method.qmd)
- `R/apc-model.R` — `prepare_apc_data()`, `build_initiation_data()`, `build_cessation_data()`
- Protocol §3.4 — analytical approach
